Finiteness in Cubical Type Theory

### 2020-06-28 sections/finiteness-predicates.tex
fixed todonotes

As a slightly more complex example, consider the \(\AgdaDatatype{Fin}\) type
we've been using.
Remember that split enumerability is in fact the same as a split surjection from
\(\AgdaDatatype{Fin}\) (Lemma~\ref{split-enum-is-split-surj}): to show that
\(\AgdaDatatype{Fin}\) is split enumerable, then, we need only show that it has
a split surjection from itself.
We'll prove the following slightly more general statement:
\begin{lemma}
  Every type has a split surjection into itself: the identity function.
\end{lemma}
\begin{proof}
  We provide the simple proof here in Agda:
  \begin{agdalisting}
    \ExecuteMetaData[agda/Function/Surjective/Properties.tex]{split-surj-ident}
  \end{agdalisting}
\end{proof}
\subsection{Decidable Equality}
Since we're working in HoTT, it's quite natural to ask how these predicates
relate to non-set types, like the circle.
\begin{definition}[\(S^1\)] \label{circle-def}
  The circle, \(S^1\), can be represented in HoTT as a higher inductive type.
  \begin{agdalisting}
    \ExecuteMetaData[agda/Snippets/Circle.tex]{circle-def}
  \end{agdalisting}
  We will use it here as an example of a non-set type, i.e. a type for which not
  all paths are equal.
  This also means that it does not have decidable equality.
\end{definition}

@@

\begin{definition}[Decidable Types]
  \begin{equation}
    \mathbf{Dec}(A) \coloneqq A \uplus \neg A
  \end{equation}
\end{definition}
\begin{definition}[Discrete Types]
  A discrete type is one with decidable equality.
  \begin{equation}
    \text{Discrete}(A) \coloneqq \Pi(x, y : A) , \mathbf{Dec}(x \equiv y)
  \end{equation}
  By Hedberg's theorem \cite{hedbergCoherenceTheoremMartinLof1998} any discrete
  type is a set.
\end{definition}

>>>>

As a slightly more complex example, consider the \(\AgdaDatatype{Fin}\) type
we've been using.
Remember that split enumerability is in fact the same as a split surjection from
\(\AgdaDatatype{Fin}\) (Lemma~\ref{split-enum-is-split-surj}): to show that
\(\AgdaDatatype{Fin}\) is split enumerable, then, we need only show that it has
a split surjection from itself.
We'll prove the following slightly more general statement:
\begin{lemma}
  Every type has a split surjection into itself: the identity function.
\end{lemma}
\begin{proof}
  We provide the simple proof here in Agda:
  \begin{agdalisting}
    \ExecuteMetaData[agda/Function/Surjective/Properties.tex]{split-surj-ident}
  \end{agdalisting}
\end{proof}
\subsection{Decidable Equality}
We have now seen some examples of types which \emph{are} split enumerable: but
what types are excluded?
Clearly infinite types like \(\AgdaDatatype{\ensuremath{\mathbb{N}}}\) aren't
allowed, but we haven't yet answered the question of non-set types.
Certainly these can be finite in some sense, but can they be split enumerable?

As our running example of a non-set type we will use the circle:
\begin{definition}[\(S^1\)] \label{circle-def}
  The circle, \(S^1\), can be represented in HoTT as a higher inductive type.
  \begin{agdalisting}
    \ExecuteMetaData[agda/Snippets/Circle.tex]{circle-def}
  \end{agdalisting}
  We will use it here as an example of a non-set type, i.e. a type for which not
  all paths are equal.
  This also means that it does not have decidable equality.
\end{definition}

Any attempt to prove that this type is split enumerable will run into trouble:
our task here, then, is to show that such a proof is impossible.
To do will remember Hedberg's theorem
\cite{hedbergCoherenceTheoremMartinLof1998}:
\begin{theorem}[Hedberg's Theorem]
  Every discrete type is a set.
\end{theorem}
With this theorem in mind, then, if we prove that every split enumerable type is
discrete we will have proven that the circle is not split enumerable, because if
it were it would be discrete and therefore a set, which we know already it is
not.

At this point we will define what we mean by \emph{discrete}.
\begin{definition}[Discrete Types]
  A discrete type is one with decidable equality.
  \begin{agdalisting}
    \ExecuteMetaData[agda/Relation/Nullary/Discrete/Base.tex]{discrete-def}
  \end{agdalisting}
\end{definition}
We also use here a data type called \AgdaDatatype{Dec}.
\begin{definition}[Decidable Types]
  If some proposition is decidable it means that it is proven true or false.
  We can encapsulate this with a data type\footnotemark:
  \begin{agdalisting}
    \ExecuteMetaData[agda/Snippets/Dec.tex]{dec-def}
  \end{agdalisting}
  A proof of type \(\AgdaDatatype{Dec}\;A\) is a proof that \(A\) is either true
  or false.
\end{definition}
\footnotetext{This is not, as it happens, the definition of
  \(\AgdaDatatype{Dec}\) you will find in Agda's standard library.
The version of the type that exists there is a good deal more complicated (for
computational reasons),
although its interface behaves}

@@
